---
title: "Gov 50 Final Project"
author: "Layla Hijjawi"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Milestone 1: Project thoughts

I am unsure of what exactly I want to explore in this project. I am potentially interested in exploring data related to racial disparities. I would also be interested into political data related to polarization. Because I live in Utah, I think it would be interesting to explore political data related to recent bills blocking gender affirming care. I'd also perhaps be interested in looking at data related to socioeconomic or racial disparities in the water crisis in Utah. 

## Milestone 2: Data Selection

Does oil revenue impact democratization or the likelihood of civil war in North African or Middle Eastern nations over time?

In this study, I plan to examine the extent to which oil as a significant source of revenue is associated with better or worse democratization or the likelihood of civil war in North Africa and the Middle East. I hypothesize that the likelihood of civil war will decrease if oil revenue is present. I believe this because extra revenue could allow governments to invest in stabilizing measures regardless of the type of regime or government. I also hypothesize that democratization will decrease further for North African and Middle Eastern nations if oil revenues are present. I believe this because research related to dependence on natural resource wealth suggests that countries with economies primarily dependent on extraction of resources like oil increases authoritarianism. 

My sample is comprised of North African and Middle Eastern countries from 1945 to 1999. My unit of analysis is a country (e.g. “Cyprus”). The explanatory variable of interest is whether a nation has greater than one-third of export revenues from fuels or not. The variable is coded = 1 for greater than one-third of export revenues from fuels and = 0 for the opposite. 

My outcome variable is whether or not civil war was occurring in a given year and polity, or the index of democratization for each nation. For civil war, the variable is coded = 1 if civil war is occurring in a given year and = 0 if it is not occurring. For polity, countries are assigned scores in a range of -10 to 10, with 10 representing a consolidated autocracy and -10 representing a consolidated autocracy. 

If I observe that civil war occurs less frequently and that polity decreases for nations with oil revenue, this would provide evidence for my hypothesis. However, if I observe that civil war occurs more frequently and that polity increases for nations with oil revenue, this would provide evidence against my thesis. 

## Milestone 3

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)

data <- read_csv("data/civilwar.csv") 
```
```{r}

# Create oil_rev variable and calculate avg polity by oil_rev

polity_data <- data |>
  mutate(
    oil_rev = if_else(Oil == 1, "Oil Dependence", "Low Oil Dependence")
  ) |>
  group_by(oil_rev, year) |>
  summarize(avg_polity = mean(polity2, na.rm = TRUE)) 

# Graph polity over time for oil dependent and non-oil dependent nations

polity_plot <- polity_data |>
  ggplot(mapping = aes(x = year, y = avg_polity, color = oil_rev)) +
  geom_smooth() +
  labs(
    x = "Year",
    y = "Average Polity Score",
    title = "Polity Over Time For Oil Dependent vs Non-Oil Dependent Countries",
    color = "Oil Dependence"
  )

polity_plot


```


```{r}

# Create oil dependence variable

war_data <- data |>
  mutate(
    oil_rev = if_else(Oil == 1, "Oil Dependence", "Low Oil Dependence")
  ) 

# Graph the proportion of active civil war over time for oil-dependent and non-oil dependent countries

war_plot <- war_data |>
  group_by(oil_rev, year) |>
  summarize(prop_war = mean(war, na.rm = TRUE)) |>
  ggplot(mapping = aes(x = year, y = prop_war, color = oil_rev)) +
  geom_smooth() +
  labs(
    x = "Year",
    y = "Proportion of Active Civil War",
    title = "Proportion of Active Civil War over Time for Oil-Dependent and Non-Oil Dependent Countries",
    color = "Oil Dependence"
  )

war_plot
```


## Milestone 4

```{r}
library(broom)

# Mutates oil_rev variable 
data <- data |>
  mutate(
    oil_rev = if_else(Oil == 1, "Oil Dependence", "Low Oil Dependence")
  )

# Creates regression with polity as outcome and oil revenue/dependence as predictor
fit_1 <- lm(polity2 ~ oil_rev, data = data)

# Creates tables of results

fit_1 |>
  broom::tidy() |>
  select(term, estimate) |>
  knitr::kable(digits = 2)
```
The intercept reveals that if a country has low oil dependence (oil_rev = 0), the estimated polity is 0.12 points. The slope or oil_revOilDependence coefficient reveals that the difference in polity between countries with oil revenue dependence versus low oil revenue dependence is -4.12 points on average. In other words, on average oil dependent nations have a polity score -4.12 points lower than countries with low oil dependence. There is a negative association between being oil dependence and polity. 

```{r}

# Creates regression with war as outcome and oil revenue/dependence as predictor
fit_2 <- lm(war ~ oil_rev, data = data)

# Creates tables of results

fit_2 |>
  broom::tidy() |>
  select(term, estimate) |>
  knitr::kable(digits = 2)
```
The intercept reveals that if a country has low oil dependence (oil_rev = 0), the estimated proportion of civil war is 0.14. The slope or oil_revOilDependence coefficient reveals that the difference in proportion of civil war between countries with oil revenue dependence versus low oil revenue dependence is -0.01. In other words, on average oil dependent nations experience 1% less civil war than counties with low oil dependence. 


